---
title: "Data Import"
output: github_document
---

```{r setup, echo = FALSE}
library(tidyverse)
library(readxl)
library(haven)
library(readr)
```

In this excercise we will be using relative path (not absolute path. Absolute path = full path name). 

* Absolute: a file or folder’s “full address” on your computer

* Relative: directions to a file or folder from your current working directory

# Import Data

## CSVs

To import a csv file, we’ll use a function from `readr` (this function is included in the tidyverse package).

```{r csv litters}
litters_data = read_csv("./Example_Data/FAS_litters.csv")
litters_data = janitor::clean_names(litters_data)
```

Using `janitor::clean_names()` to clean up variable names after importing data. Doing so will take whatever the column names are and convert them to lower snake case.

The `package::function` syntax lets you use a function from a package without loading the whole library. That’s really helpful, because some packages have functions with the same name (e.g. `dplyr::filter` and `stats::filter`), and R has to choose which one you mean. In general, only load the packages you need to prevent this kind of confusion. 

#### Looking at the data

The first thing to do after importing the data (unless `read_csv` gives warnings) is to look at it. If there are unexpected results during data import, you’ll catch a lot of them here. In addition to printing the data, I often use `View`/`view` (use in console because they don't work well in rmd files), `str`, `head`, and `tail`:

```{r look at data}
litters_data
head(litters_data, 5)
tail(litters_data, 5)
```

The code chunk below will show a neat summary of the dataset:
```{r data summary}
skimr::skim(litters_data)
```

#### `read_csv` options ...

In the best case, the data are stored in the csv file without any weirdness – there are no blank lines or columns, the first row is the variable name, missing values are stored in sensible ways. 
When this isn’t the case, arguments to read_csv are helpful.

Examples:

* `col_names`: Is the first row in the your dataset the variable names? By default, it is `TRUE`. If `FALSE`, column names are `X1`, `X1`, … . You can also supply column names.

* `na`: Telling R what values you would want to treat as 'missing values'. By default, R will treat empty rows and `NA` as missing values, but you may want to add other values (eg. 999, na, . , etc.).

* `skip`: number of rows to skip before reading data.

```{r exclude rows, eval = FALSE}
litters_data = read_csv("Example_Data/FAS_litters.csv", na = c("", "NA", 999, 88,) skip = 2)
```

###### `col_types`

`The read_*` functions will attempt to guess the data type stored in each column. By default, these guesses are based on the first 1000 rows. The guesses are also usually pretty good. In some cases, though, you’ll want to give explicit column specifications (eg. maybe R guess the variable type wrong, and you want to fix it). This is done using the `cols` function, and each column is given a column type:

```{r define column type, EVAL = FALSE}
litters_data = read_csv(file = "./Example_data/FAS_litters.csv",
  col_types = cols(
    Group = col_character(),
    `Litter Number` = col_character(),
    `GD0 weight` = col_double(),
    `GD18 weight` = col_double(),
    `GD of Birth` = col_integer(),
    `Pups born alive` = col_integer(),
    `Pups dead @ birth` = col_integer(),
    `Pups survive` = col_integer()
  )
)
```

An easier way to do this:
```{r}
litters_data = read_csv(file = "./Example_Data/FAS_litters.csv",
  col_types = "ccddiiii"
)
```

## Excel

`read_csv` is built in with RStudio, but `read_excel` is not. That's why we need to load the package `readxl` at the beginning.
If we need more information with `read_excel`, type `?read_excel` in the console.

```{r read excel spreadsheet}
mlb_df = read_excel("Example_Data/mlb11.xlsx")
```

Next step, check and see if your dataset looks good.
```{r view data, eval = FALSE}
view(mlb_df)
```

##### `range` argument for Excel
```{r range}
lotr = read_excel(
  "Example_Data/LotR_Words.xlsx",
  range = "B3:D6"
)
```

## SAS

`haven`is used to import into R data files from SAS, Stata, and SPSS. `haven` is not built in with `tidyverse`, so we need to load it at the beginning.

```{r}
pulse_df = read_sas("Example_data/public_pulse_data.sas7bdat")

head(pulse_df, 5)
```

# Export Data

All the code chunks above only allow us to read different types of datafile in R, but it will not be saved in a folder. Let's say we edited a datafile in a way that we want it to present. If we want that dataset to be saved in a folder, we need to export it.

```{r save dataset}
write_csv(lotr, file = "Results/lotr.csv")
```
The code chunk above indicates that I want to save a csv file(name of the dataset you want to save, save it to a path)


## Why not Base R???

Check this out.
```{r dont do this}
dont_do_this_df = read.csv("Example_Data/FAS_litters.csv")
```
`read.csv` is old version (Base R). It will import csv as data frame, and it will show us the entire data frame (imagine when we have a huge dataset..thousands of rows..) because R prints out the entire data frame. Sometimes it will make the character vector a factor vector, and we probably don't want that.

`read_csv` imports csv as tibbles but not data frame.



